---
title: "class08"
author: "Erica Yue Hu (PID: A17787225)"
format: pdf
---

## background 

in today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breat cancer biopsy data. 

## data import

the data is in csv format 
```{r}
read.csv("WisconsinCancer.csv")
```


```{r}
fna.data <- "WisconsinCancer.csv"

wisc.df <- read.csv(fna.data, row.names=1)

wisc.df
```

>row.names =1 means that the first column are row names not regular data column 

wee peak at the data 
```{r}
head(wisc.df, 3)
```
>Q1. how many observations are in this datatset? 

569 
```{r}
nrow(wisc.df)
```

>Q2. how many observations have a malignant diagonsis 

212

```{r}
sum(wisc.df$diagnosis == "M")
table(wisc.df$diagnosis)
```

>Q3. how many variables/features in the data are suffixed with _mean? 

```{r}
length(grep("_mean", colnames(wisc.df)))
```
10 features. 
the pattern for grep is grep(pattern, x). pattern is what you want, x is where you want it from. 

remove first column "diagnosis" from the data set cuz we don't wanna pass this to PCA. we will save this as a separate vector that we can useater to compare our findings to those of exprert. 
```{r}
wisc.data <- wisc.df[,-1]
diagnosis <-wisc.df$diagnosis
```

## principal compoent analysis (pca)

the main function in base r is called `prcomp()`. we will use the optional arugement `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the original data set. 

prcomp(x, scale = F, center = T)

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```
```{r}
wisc.pr$x
```

```{r}
library(ggplot2) 

ggplot(wisc.pr$x)+
  aes(PC1, PC2, col=diagnosis)+
  geom_point()
```


```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data,scale=T)
summary(wisc.pr)
```
>Q. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

Proportion of variance for PC1 = 0.4427

>Q. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

first 3 


> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PCA1 TO PCA7 


## interpreting pca 

```{r}
biplot(wisc.pr)
```
>Q. What stands out to you about this plot? Is it easy or difficult to understand? Why?


the only thing that stands out is the diff in colors but other than that, pretty confusing 


```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(x=PC1 , y=PC2, col=diagnosis) +
  geom_point()

```

>Q. generate another one but for components 1 and 3, what do you notice?

i noticed that the colors are the sam but its just lower on the graph 

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components:

```{r}
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

```

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
abline(h=10, col="red", lty=2)


```



```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```


## combining methods 

the idea here is that i can take my new variable (the PCs `wis.pr$x`) that are better descriptors of the data-set than the original features(i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering. 


```{r}
pc.dist<-dist(wisc.pr$x[ , 1:3]) 
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust,k=2)
table(grps)
```
i can now run `table()` with both my clustering `grps` and the expert `diagnosis` 
```{r}
table(grps, diagnosis)
```
Our cluster "1" has 179  "M" diagnosis 
Our cluster "2" has 333 "B" diagonsis. 

TP (true positive) - there are 179 true positives 
FP (false positives)- there are 24 false positives 
TN (true negatives) - there are 333 true negatives 
FN (false negatives) - there are 33 false negatives 

sensitivity: TP/(TP+FN) 

```{r}
179/(179+33)
```

specificity: TN/(TN+FP)

```{r}
333/(333+24)
```

## prediction 

we can use our pca model for prediction of new un-seen cases

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q16. prioritize patient 2
