---
title: "class 12"
author: "Erica Yue Hu (PID: A17787225)"
format: pdf
---

# section 1. proportion og G/G in a population 

downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:43440445-43443746;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel> 

here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)

```
now let's look at a different population. I picked the GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

find proportion of G|G 

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```


This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Lets now dig into this further. 


## section 4: population scale analysis 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

how mnay samples do we have? 
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head (expr)
```
```{r}
nrow(expr)
```
>. Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```


>Q14: 

let's make a boxplot 

```{r}
ggplot(expr) + aes(geno, exp,fill=geno)+
  geom_boxplot(notch=TRUE)
```




