---
title: "Class 5: Data Viz with ggplot"
author: "erica (pid:a17787225)"
format: pdf
toc: true 
---

## Background

there are lot's of ways to make plots in R i.e lots of graphics in R. These include so called "base R" (like the `plot()`) and add on packages like **ggplot2**

Let's make the same plot with these two graphics system. We can use the inbuilt `cars` dataset: 

```{r}
head( cars )
```

With "base R" we can simply: 

```{r}
plot(cars)
```
Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`

> **N.B.** we never run an `install.packages()` in a code chunk otherwise we will re-install needlessly everytime we render our document. 

every time we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```


every ggplot needs at least 3 things: 

1. The **data** i.e. stuff to plot as a data.frame 
2. The **aes** or aesthetics that map the data to the plot 
3. The **geom_** or geometry i.e the plot tyep such as points, lines etc. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist)+
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x="Speed (MPH)",
       y="Distance (ft)", 
       title="Stopping Distance of Old Cars")+
  theme_bw()

```


## Gene Expression Plot

read some data on the effects of GLP-1 inhibitor (drug) on gene expression values:
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page. 

```{r}
ggplot(genes)+
  aes(Condition1, Condition2) +
  geom_point(col="blue",alpha=0.2)
```
Let's color by `State` up, down or no change 
```{r}
table(genes$State)

```
```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State) +
  geom_point()+
  scale_colour_manual( values=c("purple","gray","pink") )+
  labs(x="control (no drugs)",
       y="drugs", 
       title="expression changes upon administeration of GLP-1 drug")+
  theme_minimal()

```

## going further with gapmider 

here we explore the famous `gampmider` dataset with some custom plots 
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. how many rows does this dataset have? 

```{r}
nrow(gapminder)
```

> How many different continents are in this dataset? 

```{r}
table(gapminder$continent)
```
Version 1 plot gdpPercap vs LifeExp for all rows 
```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+ 
  labs(x="GDP per capita",
       y="Life Expectancy",
       title = "GDP per Capita vs Life Expectancy")
```

I want to see a plot for each continent - in ggplot lingo this is called "faceting" 

```{r}
ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+ 
  facet_wrap(~continent)
  labs(x="GDP per capita",
       y="Life Expectancy",
       title = "GDP per Capita vs Life Expectancy")

```
## First look at the dplyr package 

another add-on package with a function called `filter()` that we want to use. 

```{r}
library(dplyr)
```
```{r}
filter(gapminder, year==2007, country =="United States")
```
```{r}
input <- filter(gapminder, year == 2007 | year == 1977)
```
"|" means "or" 
```{r}
input <- filter(gapminder, year == 2007 | year == 1977)
ggplot(input)+
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+ 
  facet_wrap(~year)
  labs(x="GDP per capita",
       y="Life Expectancy",
       title = "GDP per Capita vs Life Expectancy")
```

