---
title: "Class 18"
author: "Erica Yue Hu (A17787225)"
format: html
---
## background 
pertussis is a common lung infection caused by the bacteria b.pertussis. 

this can affect people of any age, but affect the most for infants 

the cdc track the number of reported ases in US (https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html): 

We can "scarpe" this data with the **datapasta** package 

```{r}
cdc <-  data.frame(
                                  Year = c(1922L,1923L,1924L,1925L,
                                           1926L,1927L,1928L,1929L,1930L,
                                           1931L,1932L,1933L,1934L,1935L,
                                           1936L,1937L,1938L,1939L,1940L,1941L,
                                           1942L,1943L,1944L,1945L,1946L,
                                           1947L,1948L,1949L,1950L,1951L,
                                           1952L,1953L,1954L,1955L,1956L,1957L,
                                           1958L,1959L,1960L,1961L,1962L,
                                           1963L,1964L,1965L,1966L,1967L,
                                           1968L,1969L,1970L,1971L,1972L,
                                           1973L,1974L,1975L,1976L,1977L,1978L,
                                           1979L,1980L,1981L,1982L,1983L,
                                           1984L,1985L,1986L,1987L,1988L,
                                           1989L,1990L,1991L,1992L,1993L,1994L,
                                           1995L,1996L,1997L,1998L,1999L,
                                           2000L,2001L,2002L,2003L,2004L,
                                           2005L,2006L,2007L,2008L,2009L,
                                           2010L,2011L,2012L,2013L,2014L,2015L,
                                           2016L,2017L,2018L,2019L,2020L,
                                           2021L,2022L,2023L,2024L,2025L),
          No..Reported.Pertussis.Cases = c(107473,164191,165418,
                                           152003,202210,181411,161799,197371,
                                           166914,172559,215343,179135,265269,
                                           180518,147237,214652,227319,
                                           103188,183866,222202,191383,191890,
                                           109873,133792,109860,156517,74715,
                                           69479,120718,68687,45030,37129,
                                           60886,62786,31732,28295,32148,
                                           40005,14809,11468,17749,17135,
                                           13005,6799,7717,9718,4810,3285,
                                           4249,3036,3287,1759,2402,1738,1010,
                                           2177,2063,1623,1730,1248,1895,
                                           2463,2276,3589,4195,2823,3450,
                                           4157,4570,2719,4083,6586,4617,
                                           5137,7796,6564,7405,7298,7867,
                                           7580,9771,11647,25827,25616,15632,
                                           10454,13278,16858,27550,18719,
                                           48277,28639,32971,20762,17972,
                                           18975,15609,18617,6124,2116,3044,
                                           7063,22538,21996)
        )
```

>Q1: make a plot of year vs cases: 

```{r}
library(ggplot2)
ggplot(cdc,aes(Year, No..Reported.Pertussis.Cases))+
  geom_point()+
  geom_line()
```
>Q. add some major milestones including the first wP vaccine roll-out (1946), the switch to the newer aP vaccine (1996), the COVID years (2020). 

```{r}
library(ggplot2)
ggplot(cdc,aes(Year, No..Reported.Pertussis.Cases))+
  geom_point()+
  geom_line()+
  geom_vline(xintercept=1946, col="blue", lty=2)+
  geom_vline(xintercept=1996, col="pink", lty=2)+
  geom_vline(xintercept = 2020, col="orange", lty=2)
```
there were high case numbers in the pre-1940s.then the number of cases decreases until around 1996, where it then started rising in cases. change from wP vaccine to aP vaccine. and it is also aP vaccine in adolescence 

key question to address: aP induced protection wanes faster than wP - why? 

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.


## CMI PB project 
computational models of immunity (https://www.cmi-pb.org/) - pertubisis boost project aims to provide the scientific community with this very information 

They make there data avaliable via JSON format APIs. We can read this in R with the `read_json()` function from the **jsonlite** package: 

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject",
                     simplifyVector = TRUE)
head(subject)
```

>Q. How many wP and aP indiviudals are in this `subject` table? 

```{r}
table (subject$infancy_vac)
```
>Q. what is the biological sex breakdown? 

```{r}
table(subject$biological_sex)
```

>Q. in terms of race and gender is this dataset representative of the US population? 

```{r}
table(subject$race, subject$biological_sex)
```

Let's read some more dataset tables 

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)

head(subject)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

to analyse this data, we need to first "join"/merge/link the different tables so we all have the data in one place, not spread across different tables. 

we can use the `*_join()` family of functions from **dplyr** to do this

```{r}
library(dplyr)
meta <- inner_join(subject, specimen) 
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

>Q. What Antibody isotypes are measured for these patients? 

```{r}
table(abdata$isotype)
```


>Q. what antigens are reported? 

```{r}
table(abdata$antigen)
```
Let's focus on the IgG isotype and make a plot of MFI_normalized for all antigens

```{r}
igg <- abdata |> 
  filter (isotype == "IgG")

head (igg)
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

>Q. is there a difference for aP vs wP individuals with these values? 

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```
>Q. is there a temprol response - i.e do values increase or decrease over time? 

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~visit)
```

## focus on "PT" pertusisis Toxin antigen

```{r}
pt.igg.21 <- igg |> filter (antigen == "PT",
               dataset == "2021_dataset")
```

```{r}
ggplot(pt.igg.21)+
  aes(planned_day_relative_to_boost, MFI_normalised,
      col = infancy_vac,
      group = subject_id)+
  geom_point()+
  geom_line()+
  geom_vline(xintercept = 14, col ="black", lty = 2)+
  geom_smooth(aes(group=infancy_vac),
              method = "loess",
              span = 1,
              se = FALSE,
              linewidth = 1.5
)
```

