---
title: "class06: r functions"
author: "erica yue hu (pid:a17787225)"
format: pdf
---

## background 

functions are at the heart of using R. Everything we do involves calling and using functions (from data input)

all functions in R have at least 3 things: 

1. A **name** the thing we use to call the function. 
2. one or more input **arguments** that are comma separated 
3. the **body**, lines of code between curly brackets { } that does the work of the function

## A first function
lets write a silly wee function to add some numbers: 
```{r}
add<- function (x) {
  x+1
}
```
lets try it out 
```{r}
add(100)
```
will ts work? 
```{r}
add(c(100,200,300))
```
modify it to be more useful and add more than just 1 
```{r}
add <- function (x,y=1) {
  x+y 
}
```
^this means that if you don't have y, it is gonna just be 1 
```{r}
add(100,10)
```
```{r}
add (100)
```
```{r}
plot(1:10, col="blue",typ="b")
```
```{r}
log(10, base = 10)
```
>**N.B** Input arguments can be either **required** or **optional**. the later have a fall-back default that is specifed in the function code with an equals sign.

```{r}
#add(x=100,y=200,z=300) ts does not work cuz we have notin for z. 
```

##A second function 
all functions in R look like this 
```
name <- function (arg) {
  body
}
```
the `sample(0)` function in R takes a sample of the specified size from the elements of x using either with or without replacement.
```{r}
sample(1:10, size = 4)

```


>Q. Return 12 numbers picked randomly from the input 1:10 

```{r}
sample(1:10, size =12, replace = TRUE)
```
`replace = true` allows you to get the same number for more than once 


>Q. write the code to generate a random 12 nucleotide long DNA sequence? 

```{r}
sample( c("A","C","G","T"), size = 12, replace = TRUE)
```
another way to write ts: 
```{r}
bases <-  c("A","C","G","T")
sample(bases, size=12, replace = TRUE)
```

>Q. write a first version function called `generate_dna()` that genarates a user specified length `n` random DNA sequence? 

``` 
name <- function (arg) {
  body
} 
```

```{r}
generate_dna <- function(n=6) {
  bases <- c("A","C","G","T")
  sample(bases, size = n, replace = TRUE)
}
```
```{r}
generate_dna(100)
#this calls the function 
```
n = 6, just means tht if you don't put anything in, then it is gonna run as 6 

> Q.modify ur function to return a FASTA like seequence so rather than [1] "G" "C" "A" "A" "T" we want "GCAAT". 

```{r}
generate_dna <- function(n=6) {
  bases <- c("A","C","G","T")
  seq=sample(bases, size = n, replace = TRUE)
  paste(seq, collapse = "")

}
```
```{r}
generate_dna(10)
```

>Q. give the user an option to return FASTA format output sequence or strandard multi-element vector format 

```{r}

generate_dna <- function(n=6, fasta=TRUE) {
  bases <- c("A","C","G","T")
  seq<-sample(bases, size = n, replace = TRUE)
  
  if(fasta){
    seq<-paste(seq, collapse="")
    cat("Hello...")
  } else { 
    cat ("...is it me ur looking for...")
  }

  
    return(seq)

}
```

```{r}
generate_dna(10,fasta=T)
generate_dna(10,fasta=F)
```
if fasta condition is fulfilled, then it is going to do everything in the sequence that is listed inside the if loop 


##A new cool function 
>Q. write a funciton called `generate_protein()` that generates a user specified length protein sequence in FASTA like format? 

>Q. use ur new `generate_protein()` function to generate  sequences between length 6 and 12 amino acids in legnth and check any of these are unique in nature (i.e found in the NR database at NCBI)


```{r}
generate_protein<- function(n=10){
  aa <- c("A","R","N","D","C","Q","E","G","H","I",
  "L","K","M","F","P","S","T","W","Y","V",
  "U")
  seq = sample(aa, size=n, replace=TRUE)

  return (paste(seq, collapse=""))
    
}
```
```{r}
generate_protein(30)
```
>Question 2: using a for loop 


```{r}
generate_protein<- function(n){
  aa <- c("A","R","N","D","C","Q","E","G","H","I",
  "L","K","M","F","P","S","T","W","Y","V",
  "U")
  seq = sample(aa, size=n, replace=TRUE)
  
  return (paste(seq, collapse=""))
}
```
```{r}
for (i in 6:12) {
  cat(">", i, "\n", sep = "")
  cat (generate_protein(i),"\n")
}


```













