---
title: "Class 7: Machine Learning 1"
author: "Erica Yue Hu (PID:A17787225)"
format: pdf
---

##Background 
today we will begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**

Let's make up some input data where we know what the natural groupings are 

The function `rnorm()`can be useful here.
```{r}
hist(rnorm(500, mean=10))

```
>Q. generate 30 random numbers centered at +3 and another 30 centered at -3


```{r}
tmp<-c(rnorm(30, mean=3), (rnorm(30,mean=-3)))

cbind(x=tmp, y=rev(tmp))
x<-cbind(x=tmp, y=rev(tmp))
plot(x)

```
rev(): reverse 
cbind(): binds them as coloumns tgt 
rbind(): binds them as rows 

##Kmeans clustering

the main function in "base R" for K-means clustering is called `kmeans()`

```{r}
km <- kmeans(x, centers=2)
km
```
>Q. what component of the results object details the clusters sizes 


```{r}
km$size
```
>Q. what component of the results object details the cluster centers? 

```{r}
km$centers
```
>what component of the results object details the cluster membership vector(i.e our main result of which points lie in which cluster) 

```{r}
km$cluster
```

>Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points 

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15)
```
>Q. Run 'kmeans()' again and this time produce 4 clusters *and call your result object `k4`) and make a results figure like above? 

```{r}
k4 <- kmeans(x, center=4)
k4
plot(x, col=k4$cluster)
points(k4$centers,col="blue", pch=15)
```


the metric 
```{r}
km$tot.withinss
k4$tot.withinss
```

>Q. let's try different number of K(centers) from 1 to 30 and see what the best result is? 

```{r}
i <- 1
ans <- NULL 
for (i in 1:30) {
  ans <- c(ans, kmeans(x,centers = i)$tot.withinss)
}
ans 
```
```{r}
plot(ans,typ="o")
```
> the best result is the point after the drop 

**key points**: k means will impose a clustering structure on your data even if it is not there - it will always give you the answer you asked for even if that answer is silly! 

##hierachical clustering 

the main function for hierachical clustering is called `hclust()`. Unlike `kmeans()` (which does all the work for you) you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function. 


```{r}
d<-dist(x)
hc <- hclust (d)
hc
plot(hc)
```

To extract our cluster membership vector from a `hclust()` result obect we have to "cut" our tree at a given height to yield a separate "groups"/"branches". 

```{r}
plot(hc)
abline(h=8, col="red", lyt=2)
```

To do this, we use the `cutree()` function on our `hclust()` object: 
```{r}
grps <- cutree(hc, h=8)
grps
```
```{r}
table(grps,km$cluster)
```
## PCA of UK food data

import the dataset of food consumption in the UK 


```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
>Q1. how many rows and columns are in your new data frame named x? what r functions could you use to answer this question? 

```{r}
dim(x)
```

one solution to set the row names is to do it by hand... 
```{r}
#rownames(x) 
rownames(x) <- x[,1]
rownames(x)
```
to remove the first column, i can use the minus index trick 

```{r}
x <- x[,-1]
x
```


A better way to do this is to set the row names to the first column with `read.csv()`

```{r}
x <- read.csv(url, row.names=1)
x
```

>Q2. which approach to solving the "row-names problem" mentioned above do you prefer and why? is one approach more robust than another under certain cricumstances? 

prefer the 2nd method because it doesn't take away a column each time you run the code. also it is shorter. 

##spotting major differens and trends 
is diffuclt even in this wee 17D dataset 
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
```{r}
barplot(as.matrix(x), beside = F, col=rainbow(nrow(x)))
```

>Q. changing what optional argument in the above barplot() function results in the following plot? 

changing the "beside" argument from T to F changes it from a histogram graph to a stacked graph. 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

```{r}
library(pheatmap)
pheatmap(as.matrix(x))
```

## PCA to the rescue 

the main pca function in "base R" is called `prcomp() `. this function wants the transpose of our food data as input (i.e. the foods as columns and the countries as rows)
```{r}
pca <-prcomp(t(x))
```
```{r}
summary(pca)
```


```{r}
attributes (pca)
```

To make one of main PCA result figures we turn to 'pca$x' the scores along our new PCs. This is called "PC plot" or"score plot" or "Ordination plot"...



```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
```

```{r}
library(ggplot2)

ggplot(pca$x)+
  aes(PC1,PC2)+
  geom_point(col=my_cols)
```

the second major result rigure is called a "loadings plot" of "variable contributions plot" or "weight plot". 
```{r}
ggplot(pca$rotation)+ 
  aes(PC1,rownames(pca$rotation))+
      geom_col()
```

